---
title: "MMD VAE rep Analysis"
author: "William Casazza"
date: "May 31, 2019"
output: beamer_presentation
---

```{r setup, include=FALSE}
library(tidyverse)
library(scales)
setwd("~/vaecit/analyses")
knitr::opts_chunk$set(echo = TRUE)
```
# Breaking down performance of autoencoder vs PCA vs other dimensionality reduction methods
There are a few experiments here that I want to cover:
1. Variability in autoencoder mediation called due to different length training
2. What type of mediation events are called by LFA, Kernel PCA, and FastICA
3. How correlated PCs and autoencoder LVs call mediation differently opposed to LFA, Kernel PCA and FastICA

Here I'll be loading in data into a clean dataframe (the previous analyses were starting to get messy):
```{r}

cit_df <- read.table("../data/rosmap_cit_pca/CIT_groupby_order.txt", header = TRUE) %>% select(snp,probes,peaks,gene,pCausal,pReact) %>% rename(pub_omni_p=pCausal,rev_pub_omni_p=pReact)
cit_rep_1_df <- read.csv("../data/rosmap_complete_rep/perm_test_cit_sanity_check_rep_1.csv") %>%
  rename(rep_1_p1 = p1, rep_1_p2 = p2, rep_1_p3 = p3, rep_1_p4 = p4, rep_1_omni_p = omni_p) %>% mutate(rep_1_part_p = pmax(rep_1_p1,rep_1_p2,rep_1_p3))
cit_rev_rep_1_df <- read.csv("../data/rosmap_complete_rep/perm_test_rev_cit_sanity_check_rep_1.csv") %>%
  rename( rev_rep_1_p1=p1, rev_rep_1_p2=p2, rev_rep_1_p3=p3, rev_rep_1_p4=p4, rev_rep_1_omni_p=omni_p) %>% mutate(rev_rep_1_part_p = pmax(rev_rep_1_p1,rev_rep_1_p2,rev_rep_1_p3)) 
cit_rep_df <- cbind(cit_df,cit_rep_1_df,cit_rev_rep_1_df)# MAIN DF
head(cit_rep_df)

pca_cit_rep_1_df <- read.csv("../data/rosmap_cit_pca/perm_test_pca_1_latent_cit_rep_1.csv") %>%
    rename(pca_rep_1_p1 = p1,pca_rep_1_p2 = p2,pca_rep_1_p3 = p3,pca_rep_1_p4= p4,pca_rep_1_omni_p = omni_p ) %>%
    mutate(pca_rep_1_part_p = pmax(pca_rep_1_p1,pca_rep_1_p2,pca_rep_1_p3))
pca_cit_rev_rep_1_df <- read.csv("../data/rosmap_cit_pca/perm_test_rev_pca_1_latent_cit_rep_1.csv") %>%
    rename(pca_rev_rep_1_p1 = p1,pca_rev_rep_1_p2 = p2,pca_rev_rep_1_p3 = p3,pca_rev_rep_1_p4= p4,pca_rev_rep_1_omni_p = omni_p ) %>%
    mutate(pca_rev_rep_1_part_p = pmax(pca_rev_rep_1_p1,pca_rev_rep_1_p2,pca_rev_rep_1_p3))
pca_df <- cbind(pca_cit_rep_1_df,pca_cit_rev_rep_1_df)# MAIN DF
head(pca_df)

mmd_cit_rep_1_df <- read.csv("../data/rosmap_cit_mmdvae/perm_test_mmdvae_1_latent_depth_10_cit_rep_1.csv") %>%
  rename(mmd_rep_1_p1 = p1,mmd_rep_1_p2 = p2,mmd_rep_1_p3 = p3,mmd_rep_1_p4= p4,mmd_rep_1_omni_p = omni_p ) %>%
  mutate(mmd_rep_1_part_p = pmax(mmd_rep_1_p1,mmd_rep_1_p2,mmd_rep_1_p3))
mmd_cit_rev_rep_1_df <- read.csv("../data/rosmap_cit_mmdvae/perm_test_rev_mmdvae_1_latent_depth_10_cit_rep1.csv") %>%
  rename(mmd_rev_rep_1_p1 = p1,mmd_rev_rep_1_p2 = p2,mmd_rev_rep_1_p3 = p3,mmd_rev_rep_1_p4= p4,mmd_rev_rep_1_omni_p = omni_p )%>%
  mutate(mmd_rev_rep_1_part_p = pmax(mmd_rev_rep_1_p1,mmd_rev_rep_1_p2,mmd_rev_rep_1_p3))
mmd_df <- cbind(mmd_cit_rep_1_df,mmd_cit_rev_rep_1_df)# MAIN DF
head(mmd_df)

mmd_train_list <- list()
for(f in dir("../data/mmd_vae_new_train/", full.names = TRUE)){
  num_epochs<-gsub(".*/perm_test.*_mmdvae_1_latent_depth_10_(.*)_epochs.csv$","\\1",f)
  if(!grepl("epochs",f)){
    model <- paste0(gsub(".*//perm_test_(.*)_1_latent_depth_5_cit.csv$","\\1",f),"_")
  }else{
    model <- paste0("mmdvae", num_epochs,"_")
    if(grepl("rev",f)){
      model <- paste0("rev_", model)
    }
  }
  
  df <- read.csv(f) %>% mutate(part_p=pmax(p1,p2,p3)) %>% rename_all(function(x) paste0(model,x))
  mmd_train_list[[f]] <- df
}
mmd_train_df <- bind_cols(mmd_train_list) # MAIN DF
head(mmd_train_df)

alt_lv_list <- list()
for(f in dir("../data/alt_lv/", full.names = TRUE)){
  model <- gsub(".*/perm_test.*_(.*)_1_latent_cit.csv$","\\1",f)
  if(grepl("rev",f)){
    model <- paste0("rev_", model)
  }
  df <- read.csv(f)%>% mutate(part_p=pmax(p1,p2,p3)) %>% rename_all(function(x) paste0(model,"_",x))
  alt_lv_list[[f]] <- df
}
alt_lv_df <- bind_cols(alt_lv_list) # MAIN DF
alt_lv_df %>% select(contains("lfa_omni_p"))
```
## Variability in autoencoder mediation results
I retrained the autoencoder using a lower number of epochs, 20, 50, and 100
```{r}
multisnp_n <- nrow(cit_df %>% select(probes,peaks,gene) %>% unique())
classify_p_value <- function(x,y,n){
      ifelse(x < (0.05 / n) & y > (0.05 / n),
        "Partial Mediation",
        ifelse(x > (0.05 / n) & y < (0.05/ n),
          "Partial Mediation",
          ifelse(x > (0.05 / n) & y > (0.05/ n),
            "Independent Association",
            "Partial Mediation")))
    
}
mmd_mediation_df <- cbind(cit_df %>% select(probes,peaks,gene),mmd_train_df, mmd_df) %>% group_by(probes,peaks,gene)%>% summarize_all(max) %>% select(contains("part_p")) %>%
  mutate(
    mmd_original_model = classify_p_value(mmd_rep_1_part_p,mmd_rev_rep_1_part_p, multisnp_n),
    mmd_20_model = classify_p_value(mmdvaecit_20_part_p,rev_mmdvaecit_20_part_p, multisnp_n),
    mmd_50_model = classify_p_value(mmdvaecit_50_part_p,rev_mmdvaecit_50_part_p, multisnp_n),
    mmd_100_model = classify_p_value(mmdvaecit_100_part_p,rev_mmdvaecit_100_part_p, multisnp_n),
    ae_model = classify_p_value(ae_part_p,rev_ae_part_p,multisnp_n),
    ae_batch_model = classify_p_value(ae_batch_part_p,rev_ae_batch_part_p, multisnp_n),
    mmdvae_warmup_model = classify_p_value(mmdvae_warmup_part_p,rev_mmdvae_warmup_part_p,multisnp_n),
    mmdvae_batch_model = classify_p_value(mmdvae_batch_part_p, rev_mmdvae_batch_part_p, multisnp_n),
    mmdvae_batch_warmup_model = classify_p_value(mmdvae_batch_warmup_part_p, mmdvae_batch_warmup_part_p, multisnp_n)
  )
mmd_mediation_df %>% gather("test","mediation_model",ends_with("model")) %>%
  group_by(test, mediation_model) %>% 
  summarize(perc = n()/ nrow(mmd_mediation_df)) %>%
  ggplot(aes(x = mediation_model,y=perc,label=percent(perc,accuracy = 0.001),fill=mediation_model)) +
    geom_bar(stat="identity", position = position_dodge()) +
    geom_text(position = position_dodge(), vjust=0.5, hjust = 0.6, size = 3) + 
    facet_wrap(~test,ncol = 3) + 
    labs(x= NULL) + 
    theme_minimal() + 
    theme(axis.text.x=element_blank())
```
Okay, this seems pretty variable actually..., need to be more sure about model selection here. Look more into the literature?

## Looking into what's called by other methods
```{r}
mediation_df <- cbind(cit_df %>% select(probes,peaks,gene),pca_df,alt_lv_df) %>% group_by(probes,peaks,gene)%>% summarize_all(max) %>% unique()%>%
  mutate(
    pca_model = classify_p_value(pca_rep_1_part_p, pca_rev_rep_1_part_p, multisnp_n),
    lfa_model = classify_p_value(lfa_part_p,rev_lfa_part_p, multisnp_n),
    fastica_model = classify_p_value(fastica_part_p,rev_fastica_part_p, multisnp_n),
    kernelpca_model = classify_p_value(kernelpca_part_p,rev_kernelpca_part_p, multisnp_n)
  )
mediation_df %>% gather("test","mediation_model",ends_with("model")) %>%
  group_by(test, mediation_model) %>% 
  summarize(perc = n()/ nrow(mediation_df)) %>%
  ggplot(aes(x = mediation_model,y=perc,label=percent(perc,accuracy = 0.001),fill=mediation_model)) +
    geom_bar(stat="identity", position = position_dodge()) +
    geom_text(position = position_dodge(), vjust=0.5, hjust = 0.6, size = 3) + 
    facet_wrap(~test,ncol = 3) + 
    labs(x= NULL) + 
    theme_minimal() + 
    theme(axis.text.x=element_blank())
```
Okay, so kernel PCA calls mediation WAY more than the other two methods, and lfa calls more than pca or fastica. This supports the idea that it's more about nonlinearity in representing features. Also supports the narratife that a more constraints on the factors leads to lower detection of mediation

## In Depth Comparisons of methods
### P value comparisons
One easy thing I can do is just compare the omnibus p values between methods to see whether or not the reason I call mediation in one case vs the other is more due to  p values being at the edge of my threshold, rather than true differences in how "rare" the signal each model capures is:
```{r}
library(GGally)

# Deep methods
cbind(mmd_train_df, mmd_df) %>%
    select(contains("omni_p"), -contains("rev"), -matches("[0-9]")) %>% 
    rename_all(function(x) gsub("_omni_p","",x)) %>% 
    mutate_all(function(x) -log10(x)) %>% 
    ggpairs(
      progress = F,
      lower = list(continuous = wrap("points", size =0.1)),
      upper = list(continuous = wrap("cor", size=3)))+
    theme(text=element_text(size=8))
cbind(mmd_train_df, mmd_df) %>%
    select(matches(".*rev.*omni_p"),-matches("[0-9]")) %>%
    rename_all(function(x) gsub("_omni_p","",x)) %>% 
    mutate_all(function(x) -log10(x)) %>%
    ggpairs(
      progress = F,
      lower = list(continuous = wrap("points", size =0.1)),
      upper = list(continuous = wrap("cor", size=3)))+
    theme(text=element_text(size=8))

# Non-deep methods
cbind(pca_df, alt_lv_df) %>%
    select(contains("omni_p"), -contains("rev"), -matches("2|3")) %>% 
    rename_all(function(x) gsub("_omni_p","",x)) %>% 
    mutate_all(function(x) -log10(x)) %>% 
    ggpairs(
      progress = F,
      lower = list(continuous = wrap("points", size =0.1)),
      upper = list(continuous = wrap("cor", size=3)))+
    theme(text=element_text(size=8))

cbind(pca_df, alt_lv_df) %>%
    select(matches(".*rev.*omni_p"),-matches("2|3")) %>%
    rename_all(function(x) gsub("_omni_p","",x)) %>% 
    mutate_all(function(x) -log10(x)) %>%
    ggpairs(
      progress = F,
      lower = list(continuous = wrap("points", size =0.1)),
      upper = list(continuous = wrap("cor", size=3)))+
    theme(text=element_text(size=8))




# ALL METHODS
cbind(mmd_train_df, mmd_df,pca_df,alt_lv_df) %>%
    select(contains("omni_p"), -contains("rev"), -matches("2|3"), -matches("[0-9][0-9]")) %>%
    rename_all(function(x) gsub("_omni_p","",x)) %>% 
    mutate_all(function(x) -log10(x)) %>%
    ggpairs(
      progress = F,
      lower = list(continuous = wrap("points", size =0.1)),
      upper = list(continuous = wrap("cor", size=3)))+
    theme(text=element_text(size=8))

cbind(mmd_train_df, mmd_df,pca_df,alt_lv_df) %>%
    select(matches(".*rev.*omni_p"), -matches("2|3"), -matches("[0-9][0-9]")) %>%
    rename_all(function(x) gsub("_omni_p","",x)) %>% 
    mutate_all(function(x) -log10(x)) %>%
    ggpairs(
      progress = F,
      lower = list(continuous = wrap("points", size =0.1)),
      upper = list(continuous = wrap("cor", size=3)))+
    theme(text=element_text(size=8))

  
```
## P Checking P value distributions for each method
```{r}
cbind(pca_df,alt_lv_df,mmd_df,mmd_train_df) %>% 
  gather("model", "Omnibus_P", matches("omni_p"), -matches("rev")) %>%
  ggplot(aes(x = Omnibus_P)) + 
    geom_histogram(bins=100) + 
    facet_wrap(~model)

cbind(pca_df,alt_lv_df,mmd_df,mmd_train_df) %>% 
  gather("model", "Omnibus_P", matches(".*rev.*omni_p")) %>%
  ggplot(aes(x = Omnibus_P)) + 
    geom_histogram(bins=100) + 
    facet_wrap(~model)

# test_results <- cbind(pca_df,alt_lv_df,mmd_df,mmd_train_df) %>% 
#   mutate(
#     mmd_original_min = pmin(mmd_rep_1_omni_p,mmd_rev_rep_1_omni_p),
#     ae_min = pmin(ae_omni_p,rev_ae_omni_p,n()),
#     ae_batch_min = pmin(ae_batch_omni_p,rev_ae_batch_omni_p),
#     mmdvae_warmup_min = pmin(mmdvae_warmup_omni_p,rev_mmdvae_warmup_omni_p),
#     mmdvae_batch_min = pmin(mmdvae_batch_omni_p, rev_mmdvae_batch_omni_p),
#     mmdvae_batch_warmup_min = pmin(mmdvae_batch_warmup_omni_p, mmdvae_batch_warmup_omni_p),
#     pca_min = pmin(pca_rep_1_omni_p, pca_rev_rep_1_omni_p),
#     lfa_min = pmin(lfa_omni_p,rev_lfa_omni_p),
#     fastica_min = pmin(fastica_omni_p,rev_fastica_omni_p),
#     kernelpca_min = pmin(kernelpca_omni_p,rev_kernelpca_omni_p),
#     mmd_original_max = pmax(mmd_rep_1_omni_p,mmd_rev_rep_1_omni_p),
#     ae_max = pmax(ae_omni_p,rev_ae_omni_p),
#     ae_batch_max = pmax(ae_batch_omni_p,rev_ae_batch_omni_p),
#     mmdvae_warmup_max = pmax(mmdvae_warmup_omni_p,rev_mmdvae_warmup_omni_p),
#     mmdvae_batch_max = pmax(mmdvae_batch_omni_p, rev_mmdvae_batch_omni_p),
#     mmdvae_batch_warmup_max = pmax(mmdvae_batch_warmup_omni_p, mmdvae_batch_warmup_omni_p),
#     pca_max = pmax(pca_rep_1_omni_p, pca_rev_rep_1_omni_p),
#     lfa_max = pmax(lfa_omni_p,rev_lfa_omni_p),
#     fastica_max = pmax(fastica_omni_p,rev_fastica_omni_p),
#     kernelpca_max = pmax(kernelpca_omni_p,rev_kernelpca_omni_p)
#   )
# ggplot(test_results%>% gather("model","Min_P", contains("min")),aes(x = Min_P)) + 
#     geom_histogram(bins=100) + 
#     facet_wrap(~model)
# 
# ggplot(test_results%>% gather("model","Max_P", contains("max")),aes(x = Max_P)) + 
#     geom_histogram(bins=100) + 
#     facet_wrap(~model)

# classify_p_value <- function(x,y,n){
#       ifelse(x < (0.05 / n) & y > (0.05 / n),
#         "Epigenetic Mediation",
#         ifelse(x > (0.05 / n) & y < (0.05/ n),
#           "Transcriptional Mediation",
#           ifelse(x > (0.05 / n) & y > (0.05/ n),
#             "Independent Association",
#             "Unclassified")))
# 
# }

forward_df <- cbind(cit_df %>% select(probes,peaks,gene),pca_df,alt_lv_df,mmd_df,mmd_train_df) %>%group_by(probes,peaks,gene)%>% summarize_all(max) %>%
  gather("model", "for_p", contains("omni_p"), -contains("rev")) %>%
  select(model, probes,peaks,gene,for_p) 
reverse_df <- cbind(cit_df %>% select(probes,peaks,gene),pca_df,alt_lv_df,mmd_df,mmd_train_df) %>% group_by(probes,peaks,gene)%>% summarize_all(max) %>%
  gather("model", "rev_p", matches(".*rev.*omni_p")) %>%
  mutate(model = gsub("rev_|_rev_","_",model)) %>%
  select(probes,peaks,gene,rev_p)
all(forward_df%>%select(probes,peaks,gene) == reverse_df%>%select(probes,peaks,gene))
cbind(forward_df%>% select(-probes,-peaks,-gene), reverse_df%>% select(-probes,-peaks,-gene)) %>% filter(grepl("kernel",model)) %>% #filter(!grepl("20|50|100",model)) %>%
    ggplot(aes(x = -log10(for_p),y = -log10(rev_p))) +
      geom_point(size=0.1,alpha = 0.2) +
      geom_hline(yintercept = -log10(0.05 / nrow(mmd_df)), color="red",linetype=2)+
      geom_vline(xintercept = -log10(0.05 / nrow(mmd_df)), color = "red",linetype=2) +
      labs(x= "-log10(P_Epigenetic)", y = "-log10(P_Transcriptional)")
      # facet_wrap(~model)
all(reverse_df$model == forward_df$model)
```
it looks like the threshold actially matters quite a bit for calling independent vs mediation, perhaps not *as* much for unclassified vs other. I will summarize changes in calling by method by threshold chosen in a later section.

## Overlap in calling between methods
Here I want to show whether or not methods are replicating the same mediation events by gene/epigenetic variable (I will need to merge in the original CIT tests from the xQTL paper).

First I will just try to hack a bar plot that shows this, but it may be more beneficial to try something that won't repeat variables as much, perhaps an [upset plot](https://github.com/hms-dbmi/UpSetR)

```{r}
cur_data <- cbind(mediation_df,mmd_mediation_df) %>% select(probes,peaks,gene,contains("model"), -matches("20|50|100")) %>% unique()%>% ungroup()

test_lvl <- c("fastica_model","pca_model", "lfa_model", "kernelpca_model", "mmd_original_model", "ae_model", "ae_batch_model", "mmdvae_warmup_model", "mmdvae_batch_model", "mmdvae_batch_warmup_model")
(df <- cur_data %>% select(-contains("batch_warmup")) %>% gather("test", "mediation", contains("model")) %>% mutate(test = factor(test, levels = test_lvl)) %>%group_by(test, mediation) %>% 
  summarize(perc = n()/ nrow(mediation_df)) )%>%
  ggplot(aes(y = perc, fill = mediation, x = mediation, group=test)) + geom_bar(color = "black", stat="identity", position=position_dodge()) + scale_y_continuous(limits=c(0,1.25), breaks = seq(0,1,0.25), labels = c("0","25","50","75", "100"))  + labs(y = sprintf("Percentage of K = %d Tests", multisnp_n),fill = "Mediation Scenario", x = "") + ggtitle("Mediation with Multiple SNPs") + theme_minimal()+ theme(text = element_text(family = "Roboto"),axis.text.x = element_blank())
df %>% group_by(mediation) %>% summarise(max(perc))
plot_upset <- function(dat,cur_var){
  dat_new <- dat %>% select(contains("model")) %>% mutate_all(function(x) as.numeric(x == cur_var)) %>%rename_all(function(test)gsub("mmd_original", "mmdvae",gsub("warmup","w",gsub("batch","b",gsub("_model","",test))))) 
  UpSetR::upset(as.data.frame(dat_new),nsets = ncol(dat_new), order.by = "freq",nintersects = 5,)
}
plot_venn <- function(dat,cur_var){
  dat_new <- dat %>% ungroup() %>%  select(contains("model")) %>% mutate_all(function(x) x == cur_var) %>%rename_all(function(test)gsub("mmd_original", "mmdvae",gsub("warmup","w",gsub("batch","b",gsub("_model","",test))))) 
  plot(eulerr::venn(dat_new[,c("ae_b", "mmdvae_b", "kernelpca", "pca", "lfa")]))
}
plot_upset(cur_data, "Independent Association")
plot_upset(cur_data, "Epigenetic Mediation")
plot_upset(cur_data, "Transcriptional Mediation")
plot_upset(cur_data, "Unclassified")
plot_venn(cur_data, "Independent Association")
plot_venn(cur_data, "Epigenetic Mediation")
plot_venn(cur_data, "Transcriptional Mediation")
plot_venn(cur_data, "Unclassified")
cur_data %>% select(contains("model"), -contains("batch_warmup")) %>%
  rename_all(function(test)gsub("mmd_original", "mmdvae",gsub("warmup","w",gsub("batch","b",gsub("_model","",test))))) %>%
  gather("value","type", everything()) %>% group_by(value) %>% summarize_all(function(x) sum(grepl("Transcriptional Mediation",x))) %>% mutate(value = factor(value, levels = value[order(-type)]))%>%
  ggplot(aes(value, type)) + geom_bar(stat = "identity")+ labs(y = "Total Number of Tests Classified", x = "Model")+theme_minimal(base_family ="Roboto") + theme(axis.text.x=element_text(angle=45, hjust=1,vjust=1))
cur_data %>% select(contains("model"), -contains("batch_warmup")) %>%
  rename_all(function(test)gsub("mmd_original", "mmdvae",gsub("warmup","w",gsub("batch","b",gsub("_model","",test))))) %>%
  gather("value","type", everything()) %>% group_by(value) %>% summarize_all(function(x) sum(grepl("Epigenetic Mediation",x))) %>% mutate(value = factor(value, levels = value[order(-type)]))%>%
  ggplot(aes(value, type)) + geom_bar(stat = "identity") + labs(y = "Total Number of Tests Classified", x = "Model")+theme_minimal(base_family ="Roboto") + theme(axis.text.x=element_text(angle=45, hjust=1,vjust=1))
cur_data %>% select(contains("model"), -contains("batch_warmup")) %>%
  rename_all(function(test)gsub("mmd_original", "mmdvae",gsub("warmup","w",gsub("batch","b",gsub("_model","",test))))) %>%
  gather("value","type", everything()) %>% group_by(value) %>% summarize_all(function(x) sum(grepl("Independent Association",x))) %>% mutate(value = factor(value, levels = value[order(-type)]))%>%
  ggplot(aes(value, type)) + geom_bar(stat = "identity")+ labs(y = "Total Number of Tests Classified", x = "Model")+theme_minimal(base_family ="Roboto")+ theme(axis.text.x=element_text(angle=45, hjust=1,vjust=1))
cur_data %>% select(contains("model"), -contains("batch_warmup")) %>%
  rename_all(function(test)gsub("mmd_original", "mmdvae",gsub("warmup","w",gsub("batch","b",gsub("_model","",test))))) %>%
  gather("value","type", everything()) %>% group_by(value) %>% summarize_all(function(x) sum(grepl("Unclassified",x))) %>% mutate(value = factor(value, levels = value[order(-type)]))%>%
  ggplot(aes(value, type)) + geom_bar(stat = "identity")+ labs(y = "Total Number of Tests Classified", x = "Model")+theme_minimal(base_family ="Roboto") + theme(axis.text.x=element_text(angle=45, hjust=1,vjust=1))
```

## More sensitivity of calling mediation by thresholds
This is probably best done in a nice table showing percentages called for each threshold, sorted by method and then by threshold:
```{r}
classify_p_value <- function(x,y,n,threshold){
      ifelse(x < (threshold / n) & y > (threshold / n),
        "Epigenetic Mediation",
        ifelse(x > (threshold / n) & y < (threshold/ n),
          "Transcriptional Mediation",
          ifelse(x > (threshold / n) & y > (threshold/ n),
            "Independent Association",
            "Unclassified")
          ))
}
mediation_at_threshold <- function(cur_n,threshold){
  mediation_threshold_df <- cbind(cit_df %>% select(probes,peaks,gene),mmd_train_df, mmd_df,pca_df,alt_lv_df) %>% group_by(probes,peaks,gene)%>% summarize_all(max) %>% unique()%>%
    mutate(
      mmd_original_model = classify_p_value(mmd_rep_1_omni_p,mmd_rev_rep_1_omni_p, cur_n,threshold),
      ae_model = classify_p_value(ae_omni_p,rev_ae_omni_p,cur_n,threshold),
      ae_batch_model = classify_p_value(ae_batch_omni_p,rev_ae_batch_omni_p, cur_n,threshold),
      mmdvae_warmup_model = classify_p_value(mmdvae_warmup_omni_p,rev_mmdvae_warmup_omni_p,cur_n,threshold),
      mmdvae_batch_model = classify_p_value(mmdvae_batch_omni_p, rev_mmdvae_batch_omni_p, cur_n,threshold),
      mmdvae_batch_warmup_model = classify_p_value(mmdvae_batch_warmup_omni_p, mmdvae_batch_warmup_omni_p, cur_n,threshold),
      pca_model = classify_p_value(pca_rep_1_omni_p, pca_rev_rep_1_omni_p, cur_n,threshold),
      lfa_model = classify_p_value(lfa_omni_p,rev_lfa_omni_p, cur_n,threshold),
      fastica_model = classify_p_value(fastica_omni_p,rev_fastica_omni_p, cur_n,threshold),
      kernelpca_model = classify_p_value(kernelpca_omni_p,rev_kernelpca_omni_p, cur_n,threshold)
    )
  return(mediation_threshold_df %>% select(probes,peaks,gene,contains('model'))) 
}
plot_at_threshold <- function(threshold){
  p <- mediation_at_threshold(multisnp_n,threshold) %>% gather("test","mediation_model",ends_with("model")) %>%
    group_by(test, mediation_model) %>% 
    summarize(perc = n()/ nrow(mediation_df)) %>%
    ggplot(aes(x = mediation_model,y=perc,label=percent(perc,accuracy = 0.001),fill=mediation_model)) +
      geom_bar(stat="identity", position = position_dodge()) +
      geom_text(position = position_dodge(), vjust=0.5, hjust = 0.6, size = 3) + 
      facet_wrap(~test,ncol = 3) + 
      labs(x= NULL) + 
      theme_minimal() + 
      theme(axis.text.x=element_blank())
  print(p)
}

plot_at_threshold(0.001)
plot_at_threshold(0.01)
plot_at_threshold(0.05)
plot_at_threshold(0.10)
plot_at_threshold(0.15)
plot_at_threshold(0.20)
plot_at_threshold(0.25)
plot_at_threshold(0.50)
plot_at_threshold(0.75)
plot_at_threshold(1)



cur_data <- mediation_at_threshold(multisnp_n,0.05)
plot_upset(cur_data, "Independent Association")
plot_upset(cur_data, "Epigenetic Mediation")
plot_upset(cur_data, "Transcriptional Mediation")
plot_upset(cur_data, "Unclassified")
cur_data <- mediation_at_threshold(multisnp_n,0.20)
plot_upset(cur_data, "Independent Association")
plot_upset(cur_data, "Epigenetic Mediation")
plot_upset(cur_data, "Transcriptional Mediation")
plot_upset(cur_data, "Unclassified")
```



## Analysis of Latent Representations
First let's load in these latent reps and see how they correlate with one another for respective genes:
```{r}
lv_dfs <- list()
for(f in list.files("../data/mmd_vae_new_train_latent",full.names = T)){
  testing <- read.csv(f)
  method <- gsub(".*/(.*)_lvs_1kg_cit_replication.csv", "\\1",f)
  testing <- testing %>% rename(sample_id = X)
  lv_dfs[[method]] <- testing
}
lvs_df <- bind_rows(lv_dfs,.id = "method")
head(lvs_df)
```

```{r}
library(corrplot)
corrs <- list()
for(i in names(lv_dfs)){
  tmp<- list()
  for(gene in colnames(lv_dfs[[i]][,-c(1)])){
    
    tmp[[gene]] <- data.frame(lapply(lv_dfs, function(x) cor(x[,gene],lv_dfs[[i]][,gene])))
    
  }
  corrs[[i]] <- tmp
}
cor_df <- bind_rows(lapply(corrs,function(x)bind_rows(x,.id="gene")),.id = "method") 
```

```{r}
library(pheatmap)
to_plot <- cor_df %>%  gather("method2", "value", everything(), -method, -gene) %>% filter(method != method2) %>%  group_by(method) %>% summarize(mean_abs_cor = mean(abs(value)))
  ggplot(to_plot, aes(method, mean_abs_cor)) +
    geom_bar(stat = "identity") +
    geom_hline(yintercept = min(to_plot$mean_abs_cor), color= "red", linetype=2) + 
    theme(axis.text.x = element_text(angle = 90, hjust=1,vjust=1))
  
cor_df %>% gather("method2", "value", everything(), -method,-gene) %>% filter(method != method2) %>%
ggplot(aes(gene,method2, fill= value)) + 
  geom_tile()+
  scale_fill_gradient2(low = "dodgerblue",mid="white",high = "red") + 
  facet_wrap(~method)
annot <- cor_df %>% filter_at(vars(everything(), -method,-gene),function(x) !all(x == 1)) %>%mutate(id_col= paste0(method,gene)) %>% select(method,id_col) %>% column_to_rownames("id_col")
df <- cor_df %>% filter_at(vars(everything(), -method,-gene),function(x) !all(x == 1)) %>% mutate(id_col= paste0(method,gene)) %>% select(-method,-gene) %>% column_to_rownames("id_col")
library(RColorBrewer)
colors <- colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(100)
colors[2] <- colors[1]
colors[100] <- colors[99]
pheatmap(t(df),annotation_col = annot, show_colnames = F, legend_breaks = c(-0.99,0,0.99), color = colors)
```

FURTHER EXPLORATION GOES HERE!!!
```{r}

```

## eQTX experiment


```{r}
(eQTX_manifest <- read.table("~/eQTX_manifest.txt", sep = " ",header = T))
eQTX_list <- list()
for(f in dir("../data/eQTX_experiment/", full.names = TRUE)){
  model <- gsub(".*/perm_test.*_(.*)_1_latent_depth_5_cit.csv$","\\1",f)
  if(grepl("rev",f)){
    model <- paste0("rev_", model)
  }
  df <- read.csv(f) %>% rename_all(function(x) paste0(model,"_",x))
  eQTX_list[[f]] <- df
}
eQTX_df <- bind_cols(eQTX_list) # MAIN DF
head(eQTX_df %>% select(contains("omni_p")))
eQTX_df %>% select(contains("omni_p")) %>% gather("method", "omni_p", everything()) %>% 
  ggplot(aes(omni_p)) +
    geom_histogram() +
    facet_wrap(~method)
eQTX_df %>% select(contains("omni_p")) %>% gather("method", "omni_p", everything()) %>% group_by(method) %>% summarize(sum(is.na(omni_p)))
classify_p_value <- function(x,y,n){
      ifelse(x < (0.05 / n) & y > (0.05 / n),
        "Epigenetic Mediation",
        ifelse(x > (0.05 / n) & y < (0.05/ n),
          "Transcriptional Mediation",
          ifelse(x > (0.05 / n) & y > (0.05/ n),
            "Independent Association",
            "Unclassified")))
    
}
eQTX_mediation <- cbind(eQTX_manifest, eQTX_df) %>% group_by(gene) %>% 
    mutate(
      ae_model = classify_p_value(ae_omni_p,rev_ae_omni_p,multisnp_n),
      pca_model = classify_p_value(pca_omni_p, rev_pca_omni_p, multisnp_n),
      lfa_model = classify_p_value(lfa_omni_p,rev_lfa_omni_p, multisnp_n),
      fastica_model = classify_p_value(fastica_omni_p,rev_fastica_omni_p, multisnp_n),
      kernelpca_model = classify_p_value(kernelpca_omni_p,rev_kernelpca_omni_p, multisnp_n)
    )
eQTX_mediation %>%gather("test","mediation_model",ends_with("model")) %>% mutate(mediation_model = replace(mediation_model,is.na(mediation_model), "Unclassified")) %>% 
  group_by(test, mediation_model) %>%
  summarize(perc = n()/ nrow(eQTX_manifest)) %>%
  ggplot(aes(x = mediation_model,y=perc,group=test,label=percent(perc,accuracy = 0.001),fill=mediation_model)) +
    geom_bar(stat="identity", position = position_dodge()) +
    geom_text(position = position_dodge(), vjust=0.5, hjust = 0.6, size = 3) + 
    labs(x= NULL) + 
    theme_minimal() + 
    theme(axis.text.x=element_blank())
eQTX_mediation %>%gather("test","mediation_model",ends_with("model")) %>% mutate(mediation_model = replace(mediation_model,is.na(mediation_model), "Unclassified")) %>% 
  group_by(test, mediation_model) %>%
  summarize(perc = n()/ nrow(eQTX_manifest)) %>% filter(grepl("Mediation", mediation_model)) 


runGSEA <- function(my_title, ranks, pathway){
  ranks <- ranks_df$v
  names(ranks) <- as.character(ranks_df$gene)
  ad_results <- fgsea(pathways = gmtPathways(pathway),stats= ranks, nperm=1000)

  fgseaResTidy <- ad_results %>%
    as_tibble() %>%
    arrange(desc(NES))
  p <- ggplot(fgseaResTidy, aes(reorder(pathway, NES), NES)) +
    geom_col(aes(fill=padj<0.25)) +
    coord_flip() +
    labs(x="Pathway", y="Normalized Enrichment Score",
         title=my_title) + 
    theme_minimal()
  print(p)
  return(fgseaResTidy)
}

df <- eQTX_mediation %>%gather("test","mediation_model",ends_with("model")) %>% mutate(mediation_model = replace(mediation_model,is.na(mediation_model), "Unclassified")) %>% filter(test == "kernelpca_model", mediation_model == "Transcriptional Mediation")
```
