---
title: "PCA vs MMD VAE 2"
author: "William Casazza"
date: "March 10, 2019"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(scales)
options(warn=-1)
knitr::opts_chunk$set(echo = TRUE)
```

# Testing Dimensionality Reduction in Causal Inference (SECOND ANALYSIS)

## PCA in Place of Genotype
Step 1: Load in Data
```{r Load PCA Data}
dir()
pca_test_list <- list()
for(f in dir("../data/csvs")){
  model <- gsub("cit_(.*)_([0-9])_PCs.*.csv","\\1", f)
  num_pc <- as.numeric(gsub("cit_(.*)_([0-9])_PCs.*.csv","\\2", f))
  tmp <- read.csv(sprintf("../data/csvs/%s", f))
  tmp$model <- model
  tmp$num_pc <- num_pc
  pca_test_list[[f]] <- tmp
}
tail(pca_df <- bind_rows(pca_test_list))
```
Step 2 is pulling apart the data and plotting
### Differences in performance using different numbers of PCs
```{r Plot PCA P Values, warning=FALSE}
pca_df <- rbind_list(pca_test_list)
pca_df <- pca_df %>% mutate(Model = plyr::mapvalues(model,from = c("caus1","ind1","null"), to = c("Full Mediation", "Independent Association", "Null"))) %>% rename(nPC = num_pc)
(p <- ggplot(pca_df, aes(x=omni_p)) + 
  geom_histogram() + 
  facet_grid(model~nPC, scales = 'free_y', labeller = labeller(nPC=label_both)) + 
  labs(x = "P Value", y = "") +
  ggtitle("Omnibus Test Using Principal Components")+
  scale_y_continuous(breaks=pretty_breaks())+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1), text = element_text(face = "bold")))
ggsave("../figures/omnibus_test_pca.svg",p)

ggplot(pca_df, aes(x=p1)) + 
  geom_histogram() + 
  facet_grid(model~nPC, scales = 'free_y', labeller = labeller(nPC=label_both))+ 
  ggtitle("Test 1 P") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1))
ggplot(pca_df, aes(x=p2)) + 
  geom_histogram() + 
  facet_grid(model~nPC, scales = 'free_y', labeller = labeller(nPC=label_both))+ 
  ggtitle("Test 2 P") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1))
ggplot(pca_df, aes(x=p3)) + 
  geom_histogram() + 
  facet_grid(model~nPC, scales = 'free_y', labeller = labeller(nPC=label_both))+ 
  ggtitle("Test 3 P") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1)) 
ggplot(pca_df, aes(x=p4)) + 
  geom_histogram() + 
  facet_grid(model~nPC, scales = 'free_y', labeller = labeller(nPC=label_both))+ 
  ggtitle("Test 4 P") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1)) 
```

### Interesting Notes
Caus1 Appears to work better in the way we constructed the data, but it's important to note that it's about a 50 50 shot to call independent association, as constructed, as causal mediation. In the case where there's no association between variables we don't call causal mediation at all, which is a good behavior.


## MMD VAE

Step 1: Load in the data

```{r Load MMD VAE Data}
mmd_test_list <- list()
for(f in dir("../data/mmd_vae_tests_3/")){
  model <- gsub("cit_(.*)_mmdvae_([0-9]*)_depth_([0-9]*)_latent.*.csv","\\1", f)
  num_hidden <- as.numeric(gsub("cit_(.*)_mmdvae_([0-9]*)_depth_([0-9]*)_latent.*.csv","\\3", f))
  num_latent <- as.numeric(gsub("cit_(.*)_mmdvae_([0-9]*)_depth_([0-9]*)_latent.*.csv","\\2", f))
  tmp <- read.csv(sprintf("../data/mmd_vae_tests_3/%s", f))
  tmp$model <- model
  tmp$num_hidden <- num_hidden
  tmp$num_latent <- num_latent
  mmd_test_list[[f]] <- tmp
}
tail(mmd_df <- rbind_list(mmd_test_list))
```

Step 2 is pulling apart the data and plotting
### Differences in performance using different numbers of latent variables and hidden layers
```{r Plot MMD VAE P Values, warning=FALSE}

mmd_df <- rbind_list(mmd_test_list)
mmd_df <- mmd_df %>%  mutate(Model = plyr::mapvalues(model,from = c("caus1","ind1","null"), to = c("Full Mediation", "Independent Association", "Null"))) %>%rename(nLatent = num_latent, nHidden = num_hidden)
(p <- ggplot(mmd_df, aes(x=omni_p)) + 
  geom_histogram() + 
  facet_grid(model~nHidden + nLatent, scales = 'free_y', labeller = labeller(nHidden=label_both, nLatent = label_both)) + 
  labs(x = "P Value", y = "") +
  ggtitle("Omnibus Test Using Latent Encodings")+
  scale_y_continuous(breaks=pretty_breaks())+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1), text = element_text(face = "bold")))
ggsave("../figures/omnibus_test_mmd.svg",p)


ggplot(mmd_df, aes(x=p1)) +
  geom_histogram() +
  facet_grid(model~nHidden + nLatent, scales = 'free_y', labeller = labeller(nHidden=label_both, nLatent = label_both)) + 
  ggtitle("Causal Inference Test Condition 1 Using Latent Encodings") +
    labs(x = "P Value", y = "") +
    scale_y_continuous(breaks=pretty_breaks())+

  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1), text = element_text(face = "bold"))


ggplot(mmd_df, aes(x=p4)) +
  geom_histogram() +
  facet_grid(model~nHidden + nLatent, scales = 'free_y', labeller = labeller(nHidden=label_both, nLatent = label_both)) + 
  ggtitle("Causal Inference Test Condition 4 Using Latent Encodings") +
    labs(x = "P Value", y = "") +
  scale_y_continuous(breaks=pretty_breaks())+

  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1), text = element_text(face = "bold"))


ggplot(mmd_df, aes(x=p2)) +
  geom_histogram() +
  facet_grid(model~nHidden + nLatent, scales = 'free_y', labeller = labeller(nHidden=label_both, nLatent = label_both)) + 
  ggtitle("Test 2 P") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1))
ggplot(mmd_df, aes(x=p3)) +
  geom_histogram() +
  facet_grid(model~nHidden + nLatent, scales = 'free_y',labeller = labeller(nHidden=label_both, nLatent = label_both))+
  ggtitle("Test 3 P") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1))
# ggplot(mmd_df, aes(x=p4)) +
#   geom_histogram() +
#   facet_grid(model~nHidden + nLatent, scales = 'free_y')+
#   ggtitle("Test 4 P") +
#   theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1))
```



## VanBUG Draft plots
```{r pca VanBUG}
# condition 1 capture association
ggplot(pca_df, aes(p1, fill = Model)) + geom_histogram(position = "identity", alpha = 0.6) + scale_fill_manual(values = c("blue", "red", "gold")) + labs(y = "Count (100 simulations per model)",x = "P Value Association PC->Expression") + ggtitle("Directed Association Using Genotype PC 1")+ theme_minimal(base_family = "Arial")
ggplot(pca_df, aes(omni_p , fill = Model)) + geom_histogram(position = "identity", alpha = 0.6) + scale_fill_manual(values = c("blue", "red", "gold"))+ labs(y = "Count (100 simulations per model)",x = "P Value Epigenetic Mediation") + ggtitle("Capturing Epigenetic Mediation of Genotype PC1")+ theme_minimal(base_family = "Arial") + scale_y_continuous(limits=c(0,100))
```
```{r pca debugging}
to_plot_pca <- pca_df %>% gather(key = "test", value = "test_pvalue",p1,p2,p3,p4,omni_p)
ggplot(to_plot_pca, aes(test_pvalue, fill = Model)) + 
  geom_histogram(position = "identity", alpha = 0.6) + 
  scale_fill_manual(values = c("blue", "red", "gold")) +
  labs(y = "Count (100 simulations per model)",x = "P Value Epigenetic Mediation") +
  ggtitle("Capturing Epigenetic Mediation of Genotype PC1")+ 
  theme_minimal(base_family = "Arial") + 
  scale_y_continuous(limits=c(0,100)) + 
  facet_wrap(~test,ncol=2)
```


```{r mmdvae VanBUG}

ggplot(mmd_df, aes(p1, fill = Model)) + geom_histogram(position = "identity", alpha = 0.6) + scale_fill_manual(values = c("blue", "red", "gold")) + labs(y = "Count (200 simulations per model)",x = "P Value Association LV->Expression") + ggtitle("Directed Association with 1 Autoencoder LV")+ theme_minimal(base_family = "Arial")
ggplot(mmd_df, aes(omni_p , fill = Model)) + geom_histogram(position = "identity", alpha = 0.6) + scale_fill_manual(values = c("blue", "red", "gold")) + labs(y = "Count (200 simulations per model)",x = "P Value Epigenetic Mediation") + ggtitle("Capturing Epigenetic Mediation of 1 Autoencoder LV")+ theme_minimal(base_family = "Arial")
```
```{r mmd vae debug}
mmd_df <- mmd_df %>% mutate(p3 = 1 - p3) %>% mutate(omni_p = pmax(p1,p2,p3,p4))
to_plot_mmd <- mmd_df %>% gather(key = "test", value = "test_pvalue",p1,p2,p3,p4,omni_p)
ggplot(to_plot_mmd, aes(test_pvalue, fill = Model)) + 
  geom_histogram(position = "identity", alpha = 0.6) + 
  scale_fill_manual(values = c("blue", "red", "gold")) +
  labs(y = "Count (100 simulations per model)",x = "P Value Epigenetic Mediation") +
  ggtitle("Capturing Epigenetic Mediation of Genotype PC1")+ 
  theme_minimal(base_family = "Arial") + 
  scale_y_continuous(limits=c(0,100)) + 
  facet_wrap(~test,ncol=2)
```

```{r distribution max 4 uniform RV}
unif_tests <- matrix(runif(4000),1000,4)
ggplot(data.frame(P = apply(unif_tests,1,max)), aes(x = P)) + geom_histogram(fill="gold", alpha = 0.8)+theme_minimal()
```

